Evolution of pre-copulatory and post-copulatory strategies of inbreeding avoidance and associated polyandry

### 2017-03-18 ms_pre_post.Rmd
Edits up through Methods and first subsection of Results

@@

<!--- Avoided 'used an IBM' to prevent the redundancy of 'used an IBM to model'--->

<!--- I have added a sentence like that suggested, hitting the key evolutionary processes (as all pop-geneticists will recognise these as fundamental, emphasising that we're accounting for all the possible ways evolution can occur). It's tempting to remove 'gene-flow', which we only really use to prevent kinships from getting too high, but without it it sounds as if we're leaving something out. I don't see why this deals with the concern in the intro though -- the sentence just says what we did; it doesn't really explain why we *needed* to do what we did to address the knowledge gap. --->

@@

Phenotypic values for tendency for polyandry ($P_{p}$), pre-copulatory inbreeding strategy ($M_{p}$), and post-copulatory inbreeding strategy ($F_{p}$) each have costs that are assumed to independently increase the probability that a focal female will die before mating (see Discussion for consideration of other costs). The probabilities of pre-mating mortality due to the costs of polyandry ($c_{P}$), pre-copulatory inbreeding strategy ($c_{M}$), and post-copulatory inbreeding strategy ($c_{F}$) are $P_{p} \times c_{P}$, $|M_{p}| \times c_{M}$, and $|F_{p}| \times c_{F}$, respectively. Here $|M_{p}|$ and $|F_{p}|$ are the absolute values of $M_{p}$ and $F_{p}$, respectively. Absolute values are used for applying costs to inbreeding strategies because both negative and positive $M_{p}$ and $F_{p}$ values affect the degree of inbreeding. In contrast, only positive values of $P_{p}$ cause females to be polyandrous (see below). Overall, because generations are non-overlapping, a female's probability of total reproductive failure increases linearly with each trait value.

<!--- One issue with the 'see Discussion' though is that it's not clear what we're telling readers to see the discussion *about*. Were I to read this as reviewer, I would be confused, ''see Discussion? What for? Because you made an assumption in a model, the biological context of which was already explained in the Introduction? (2nd to last paragraph)''. I've tried to provide a bit of context in parentheses, but it makes me nervous catering to any (mistaken) readers who might be confused about what assumptions are used for in models. I'd rather deal with confused reviewers than write a paper that perpetuates some (increasingly frustrating) misunderstandings of modelling and theory.  --->

@@

Each male's value with respect to a focal female is then divided by the sum of all $Q^{m}_{ij}$ values across all males with respect to that female, thereby assigning each male a relative perceived quality $q^{m}_{ij}$, which is constrained to values between zero and one. The value of $q^{m}_{ij}$ then defines the probability that a focal female mates with the male. Mating is realised by placing each $q^{m}_{ij}$ value into a vector whose elements sum to unity. A new vector of identical size is then created where each element equals the summation of all earlier position elements in the original vector of $q^{m}_{ij}$ values (such that the first element is zero and last is a number $\leq 1$). Because each element in the vector corresponds to a unique male, a random uniform number between zero and one can identify a female's realised mate. This realisation is enacted by identifying the element in the vector that is closest to, but not less than, the random uniform number. Mating is therefore stochastic, and females do not always mate with the male of the highest $q^{m}_{ij}$. For polyandrous females that choose multiple mates (i.e., $N_{mates}>1$), mates are chosen iteratively such that $Q^{m}_{ij}$ and $q^{m}_{ij}$ are re-calculated for each additional mate choice, and with $Q^{m}_{ij}$ and therefore $q^{m}_{ij}$ values of already chosen males set to zero to ensure mate sampling without replacement. In the unlikely event that a female's $N_{males}$ exceeds the total number of available males, then she simply mates with all males.

<!--- We might want to step back a bit on this one and figure out what exactly it is that is confusing and needs clarifying, as I'm not sure I see what the issue is anymore or, rather, why we're going into so much detail about a very routine procedure in an IBM (or any bit of code, really). This kind of sampling procedure was used in both of our previous IBM papers and I don't recall it being an issue. I have explained the procedure as carefully as I can, but it's getting to the point now where it's essentially turning the specifics of the C programming language into English (i.e., the methods here are now language dependent -- could be done differently in R or Perl or whatever, essentially just to say that we're sampling an element randomly from a weighted vector (I'm still not sure why we can't go back to just saying it that way; actually, I don't see why we can't just stop after ''the value of q^m_ij then defines the probability that a focal female mates with the male'' -- why does this need further explanation in the text?)). --->

@@

<!--- I've tried another way of explaining that sampling is done from a vector of probabilities in the second to last sentence here --->

@@

The viability of a focal female $i$'s offspring ($\Psi_{\textrm{off}}$) decreases as a log-linear function of her kinship with the sire $j$ of her offspring ($k_{ij}$) and inbreeding depression slope $\beta$,
\begin{equation}
\Psi_{\textrm{off}} = e^{-\beta k_{ij}}
\end{equation}
Here, $\beta$ models the number of haploid lethal equivalents that exist as deleterious recessive alleles in the gametes of $i$ and $j$, and which might be homozygous in offspring and reduce viability. Equation 1 assumes independent allelic effects, generating multiplicative effects on offspring viability [@Morton1956; @Mills1994]. We model inbreeding depression as having an absolute rather than relative effect on offspring viability (i.e., hard rather than soft selection), so that the effect of $\beta$ is consistent across generations and different parameter combinations. We also assume that inbreeding always decreases offspring viability (i.e., $\beta > 0$, no outbreeding depression). Therefore, because $0 \leq k_{ij} \leq 1$, $-\beta \times k_{ij} \leq 0$. Values of $\Psi_{\textrm{off}}$ must therefore be between zero (if $-\beta \times k_{ij}$ is very negative) and one (if $-\beta \times k_{ij} = 0$). We therefore define $\Psi_{\textrm{off}}$ as the probability that an offspring is viable, and sample its realised viability (versus mortality) using a Bernoulli trial.

<!--- See the PDF for the equation. Not sure why it's not showing up in the DOCX. --->

@@

To address how asymmetric costs associated with pre-copulatory and post-copulatory inbreeding avoidance and polyandry affect the long-term persistence of reproductive strategies, we quantify the change in $M_{a}$ and $F_{a}$ over generations in simulations where the cost of pre-copulatory inbreeding strategy $c_{M}=0$ but the cost of post-copulatory inbreeding strategy $c_{F}=0.02$, and vice versa. We compare these simulations with evolution of a costly strategy in the absence of evolution of an alternative strategy (e.g., evolution of pre-copulatory inbreeding strategy when post-copulatory inbreeding strategy phenotype is fixed at zero, $F_{p}=0$). In Supporting Information, we present results for all possible cost combinations of $0$ and $0.02$. Previous modelling using similar genetic architecture suggests that a cost of $0.02$ imposes strong but not overhwleming direct negative selection on polyandry [@Duthie].

To address how selection on an initially rare strategy of pre-copulatory or post-copulatory inbreeding avoidance is affected by the other strategy of inbreeding avoidance already being fixed, we first used exploratory simulations to quantify evolution of pre-copulatory inbreeding strategy, and post-copulatory inbreeding strategy and associated polyandry, in isolation. Then, to test whether pre-copulatory inbreeding avoidance would evolve when adaptive polyandry and post-copulatory inbreeding avoidance were fixed, we initiated $M_{a}$ allele values at zero, but fixed $F_{a}$ allele values at $-10$ and $P_{a}$ allele values at $1$ (i.e., meaning that $F_{p}$ and $P_{p}$ did not evolve). Similarly, to test whether post-copulatory inbreeding avoidance would evolve when pre-copulatory inbreeding avoidance was already fixed, we initiated $F_{a}$ and $P_{a}$ allele values at zero but fixed $M_{a}$ allele values at $-10$. Consequently, because females have 10 diploid loci, when $M_{a}$ or $F_{a}$ alleles were fixed at $-10$, outbred females were $51$ times less likely to choose a full brother and $13.5$ times times less likely to choose a first cousin than a non-relative in pre-copulatory and post-copulatory choice, respectively.

In all simulations, we recorded mean values of $P_{a}$, $M_{a}$, and $F_{a}$ in each generation, and we present these values over generations to infer selection on phenotypes ($P_{p}$, $M_{p}$, and $F_{p}$). Each combination of parameter values simulated was replicated $40$ times, and grand mean values and standard errors of means are calculated in each generation across replicates. These analyses allowed us to infer how allele values changed over generations in response to costs, but also in response to the changing values of other alleles and therefore potential evolutionary feedbacks among reproductive strategies (e.g., selection on post-copulatory inbreeding strategy might be weak if polyandry is rare). For all replicates, we set the maximum number of generations to 40000, which exploratory simulations and previous modelling [@Duthie] showed to be sufficient for inferring long-term dynamics of mean allele values and therefore selection on phenotypes.  For all replicates, we set $\rho=5$ immigrants, which produced a range of kin and non-kin in each generation allowing females to express inbreeding strategies, and $n=8$ offspring, which was sufficient to keep populations consistently at carrying capacities and avoid population extinction. Values of $K_{f}$ and $K_{m}$ were both set to 100 because previous modelling showed that populations of this size are small enough that mate encounters between kin occur with sufficient frequency for selection on inbreeding strategy, but not so small that selection is typically overwhelmed by genetic drift [@Duthie2016a].

@@

When post-copulatory inbreeding avoidance was allowed to evolve, mean $P_{a}$ values became substantially higher than in comparable simulations where $F_{a}$ values were fixed to zero and post-copulatory inbreeding avoidance could not evolve (Figure 1B,D,F,H). Allowing evolution of post-copulatory inbreeding avoidance alongside pre-copulatory inbreeding avoidance therefore facilitated evolution of polyandry. However, $P_{a}$ values were expected to become positive only given a low cost of polyandry ($c_{P} < 0.005$, Figure 1B,D versus 1F,H). This implies that, even given strong inbreeding depression in offspring viability, inbreeding avoidance might generally be a weak force driving polyandry evolution.

<!--- I've not added the suggested ''to the degree that most females mated multiply'' because it's not actually true unless costs are less than or equal to 0.025. Also, do we really need to point out figure 2 here? The very next paragraph explains that the degree of polyandry isn't trivial, and I should hope that readers aren't so impatient or put off by Figure 1 that they stop reading entirely!  --->

Strong post-copulatory inbreeding avoidance, manifested as very negative $F_{a}$ values, consistently evolved in all simulations where such evolution was allowed (black lines, Figure 1B,D,F,H). Evolution of post-copulatory inbreeding avoidance occurred even when $P_{a}$ values were expected to be slightly negative, and hence when there was selection against polyandry, (Figure 1F,H). This reflects the threshold nature of phenotypic expression of polygenic polyandry, wherein random sampling of alleles means that polyandry is expressed expressed by some females (i.e., $P_{g}>0$) even when mean $P_{a}$ values are negative (Figure 2C,D). This means that, even in populations where female reproductive strategy evolves toward monandry, there is still commonly some opportunity for expression of post-copulatory inbreeding avoidance and associated selection that drives evolution of post-copulatory inbreeding avoidance.

<!--- I'm not sure what aspects of Figure 1 you want me to mention in the main text? Are these things that are also in the legend, and if so, do we really need to repeat them? --->

>>>>

I think people will have a problem with our approach if it is not clear why our approach is the best one for addressing our key questions. I sometimes see researchers using a population, adaptive dynamic, individual-based, etc., not because it is the appropriate model for addressing their question, but because it's the one with which they are most familiar. As a result, models can become mathematically intractable (should have used an IBM) or conceptually incoherent (should have simplified in some way). We need to be able to make it clear why we are using such a complex individual-based model instead of something else.

My concern with some unusual highlighting of assumptions isn't (of course) that we're going to make some readers less confused. My concern is that we're going to make some readers *more* confused. Theoretically literate readers might be confused by some of the assumptions we keep coming back to, thinking that we're making a biological point about their relative importance that I don't think we are. I don't want to write a paper just that gets through review by playing to people's misconceptions (even if those who hold such misconceptions are the majority) because it rienforces the idea that those misconceptions are justified. If we continue to do this, then writing and publishing good theory will get even more difficult in the future. Increasingly I fear that I am doing my colleagues who work on theory a disservice by writing papers that placating people's concerns rather than articulating why (some, but of course not all) these concerns are misplaced, and hence setting a bad precedent.

Hence, when we explain why a cost of 0.02 is appropriate, we are giving readers the impression that its value really needed to be justified. I would much prefer to explain why this isn't necessarily the case, and what a cost means in the context of exploring theory, so as to avoid reviewers of other papers point to Duthie et al. (XXXX) and say, ''they justified their cost, so you should have to as well!''. I tend to see the same thing for mutation rates -- the whole point of setting a mutation rate (usually to 0.001, for purely arbitrary reasons) is to introduce a bit of variation and see what happens. The exact rate is inconsequential with respect to most theoretical questions that models address (i.e., changing it might changes outcomes a bit, perhaps even qualitatively, but it would not change how peoples *thinking* and *conceptualisation* of biology would be affected by the model). If a reader is worried about the exact rate of mutation then they probably aren't actually understanding the point of the model.

@@

@@

Phenotypic values for tendency for polyandry ($P_{p}$), pre-copulatory inbreeding avoidance strategy ($M_{p}$), and post-copulatory inbreeding avoidance strategy ($F_{p}$) each have costs that are assumed to independently increase the probability that a focal female will die before mating (see Discussion for consideration of other costs). The probabilities of pre-mating mortality due to the costs of polyandry ($c_{P}$), pre-copulatory inbreeding avoidance strategy ($c_{M}$), and post-copulatory inbreeding avoidance strategy ($c_{F}$) are $P_{p} \times c_{P}$, $|M_{p}| \times c_{M}$, and $|F_{p}| \times c_{F}$, respectively. Here $|M_{p}|$ and $|F_{p}|$ are the absolute values of $M_{p}$ and $F_{p}$, respectively. Absolute values are used for applying costs to inbreeding avoidance strategies (hereafter simply 'inbreeding strategies') because both negative and positive $M_{p}$ and $F_{p}$ values affect the degree of inbreeding, representing inbreeding avoidance and inbreeding preference, respectively. In contrast, only positive values of $P_{p}$ cause females to be polyandrous (see below). Overall, because generations are non-overlapping, a female's probability of total reproductive failure increases linearly with the phenotypic value of each trait.

@@

Each male's value with respect to a focal female is then divided by the sum of all $Q^{m}_{ij}$ values across all males with respect to that female, thereby assigning each male a relative perceived quality $q^{m}_{ij}$, which is constrained to values between zero and one. The value of $q^{m}_{ij}$ then defines the probability that a focal female mates with the male. Mating is therefore stochastic, and females do not always mate with the male of the highest $q^{m}_{ij}$. For polyandrous females that choose multiple mates (i.e., $N_{mates}>1$), mates are chosen iteratively such that $Q^{m}_{ij}$ and $q^{m}_{ij}$ are re-calculated for each additional mate choice, and with $Q^{m}_{ij}$ and therefore $q^{m}_{ij}$ values of already chosen males set to zero to ensure mate sampling without replacement. In the unlikely event that a female's $N_{males}$ exceeds the total number of available males, then she simply mates with all males.

@@

@@

The viability of a focal female $i$'s offspring ($\Psi_{\textrm{off}}$) decreases as a log-linear function of her kinship with the sire $j$ of her offspring ($k_{ij}$) and inbreeding depression slope $\beta$,
\begin{equation}
\Psi_{\textrm{off}} = e^{-\beta k_{ij}}
\end{equation}
Here, $\beta$ models the number of haploid lethal equivalents that exist as deleterious recessive alleles in the gametes of $i$ and $j$, and which might be homozygous in offspring and reduce viability. Equation 1 assumes independent allelic effects, generating multiplicative effects on offspring viability [@Morton1956; @Mills1994]. We model inbreeding depression as having an absolute rather than relative effect on offspring viability (i.e., hard rather than soft selection), so that the effect of $\beta$ is consistent across generations and different parameter combinations. We assume that inbreeding always decreases offspring viability (i.e., $\beta > 0$, giving inbreeding depression but no outbreeding depression). Therefore, because $0 \leq k_{ij} \leq 1$, $-\beta \times k_{ij} \leq 0$. Values of $\Psi_{\textrm{off}}$ must therefore be between zero (if $-\beta \times k_{ij}$ is very negative) and one (if $-\beta \times k_{ij} = 0$). We therefore define $\Psi_{\textrm{off}}$ as the probability that an offspring is viable, and sample its realised viability (versus mortality) using a Bernoulli trial. Our model therefore considers evolution of strategies of inbreeding avoidance caused by inbreeding depression, not evolution of strategies of inbreeding preference, such as might be expected to arise given weak or zero inbreeding depression [@Parker1979; @Kokko2006; @Duthie2016a], or given outbreeding depression [@Bateson1983; @Greeff2009].

@@

To address how asymmetric costs associated with pre-copulatory and post-copulatory inbreeding avoidance and polyandry affect the long-term persistence of reproductive strategies, we quantify the change in $M_{a}$ and $F_{a}$ over generations in simulations where pre-copulatory inbreeding strategy was cost free ($c_{M}=0$) but post-copulatory inbreeding strategy was moderately costly ($c_{F}=0.02$), and vice versa. We compare evolutionary trajectories with those of a costly strategy in the absence of evolution of an alternative strategy (e.g., evolution of pre-copulatory inbreeding strategy when post-copulatory inbreeding strategy phenotype is fixed at zero, $F_{p}=0$). In Supporting Information, we present results for all possible cost combinations of $0$ and $0.02$. Previous modelling using similar genetic architecture suggests that a cost of $0.02$ imposes strong but not overwhelming direct negative selection on polyandry [@Duthie]. This value is therefore appropriate to illustrate the different evolutionary consequences that could result from asymmetrical costs.

To address how selection on an initially rare strategy of pre-copulatory or post-copulatory inbreeding avoidance, and resulting evolution, is affected by the other strategy of inbreeding avoidance already being fixed, we first used exploratory simulations to quantify evolution of pre-copulatory inbreeding strategy, and post-copulatory inbreeding strategy and associated polyandry, in isolation. Then, to test whether pre-copulatory inbreeding avoidance would evolve when adaptive polyandry and post-copulatory inbreeding avoidance were fixed, we initiated $M_{a}$ allele values at zero, but fixed $F_{a}$ allele values at $-10$ and $P_{a}$ allele values at $1$ (i.e., meaning that $F_{p}$ and $P_{p}$ did not evolve). Similarly, to test whether post-copulatory inbreeding avoidance would evolve when pre-copulatory inbreeding avoidance was already fixed, we initiated $F_{a}$ and $P_{a}$ allele values at zero but fixed $M_{a}$ allele values at $-10$. Consequently, because females have 10 diploid loci, when $M_{a}$ or $F_{a}$ alleles were fixed at $-10$, outbred females were $51$ times less likely to choose a full brother and $13.5$ times times less likely to choose a first cousin than a non-relative in pre-copulatory and post-copulatory choice, respectively.

In all simulations, we recorded mean values of $P_{a}$, $M_{a}$, and $F_{a}$ in each generation, and we present these values over generations to infer selection on phenotypes ($P_{p}$, $M_{p}$, and $F_{p}$). Each combination of parameter values simulated was replicated $40$ times, and grand mean values and standard errors of means are calculated in each generation across replicates. These analyses allowed us to infer how allele values changed over generations in response to costs, but also in response to the changing values of other alleles and therefore potential evolutionary feedbacks among reproductive strategies. For all replicates, we set the maximum number of generations to 40000, which exploratory simulations and previous modelling [@Duthie] showed to be sufficient for inferring long-term dynamics of mean allele values and therefore selection on phenotypes.  For all replicates, we set $\rho=5$ immigrants, which produced a range of kin and non-kin in each generation allowing females to express inbreeding strategies, and $n=8$ offspring, which was sufficient to keep populations consistently at carrying capacities and avoid population extinction. Values of $K_{f}$ and $K_{m}$ were both set to 100 because previous modelling showed that populations of this size are small enough that mate encounters between kin occur with sufficient frequency for selection on inbreeding strategy, but not so small that selection is typically overwhelmed by genetic drift [@Duthie2016a].

@@

When post-copulatory inbreeding avoidance was allowed to evolve, mean $P_{a}$ values became substantially higher than in comparable simulations where $F_{a}$ values were fixed to zero and post-copulatory inbreeding avoidance could not evolve (Figure 1B,D,F,H). Allowing evolution of post-copulatory inbreeding avoidance alongside pre-copulatory inbreeding avoidance therefore facilitated evolution of polyandry to the degree that most females mated multiply given low costs of polyandry ($c_{P} < 0.005$; e.g., Figure 2A,B). Hence, given a low cost of polyandry, $P_{a}$ allele values increased from zero and persisted at low positive values (Figure 1B,D). In contrast, highers cost of polyandry ($c_{P} \geq 0.005$) resulted in initial increases of $P_{a}$ allele values from zero, but then became slightly negative over generations. Trajectories of $P_{a}$ values in individual simulations were typically highly stochastic, but also consistent in long-term direction (Supporting Information pp. S5-S12). Overall, this implies that, even given strong inbreeding depression in offspring viability, inbreeding avoidance might generally be a weak force driving polyandry evolution.

Strong post-copulatory inbreeding avoidance, manifested as very negative $F_{a}$ values, consistently evolved in all simulations where such evolution was allowed (black lines, Figure 1B,D,F,H). Evolution of post-copulatory inbreeding avoidance occurred even when $P_{a}$ values were expected to be slightly negative, and hence when there was selection against polyandry (Figure 1F,H). This reflects the threshold nature of phenotypic expression of polygenic polyandry, wherein random sampling of alleles means that polyandry is expressed by some females (i.e., $P_{g}>0$) even when mean $P_{a}$ values are negative (Figure 2). This means that, even in populations where female reproductive strategy evolves toward monandry, there is still commonly some opportunity for expression of post-copulatory inbreeding avoidance and associated selection that drives evolution of post-copulatory inbreeding avoidance.
